Nonparametric Clustering from Energy Statistics

### 2017-09-17 qcqp_clustering/draft/energy_clustering.tex
more comments

\begin{abstract}
Energy statistics 
was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics,
and it provides a nonparametric test for equality of distributions.
It was generalized to 
metric spaces of strong negative type,
and more recently a connection with reproducing kernel Hilbert spaces 
was proposed.
Although  
extensively used in statistics, it
has not been previously incorporated in machine learning problems.
% @gui: not sure what you mean?  what counts as machine learning?  be specific, i think yo mean clustering?  even then, it is not true, both the master's thesis and the RKHS paper kind of do it.  be as general as possible about the gap that we are filling, but not more general.
Here we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP), which we show
to be equivalent
to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of 
kernel $k$-means from energy statistics.
% @gui: have other people shown a relationship between kernel k-means and say, RKHS, or something else?  
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, which makes connection to 
graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative
algorithm based on Hartigan's method, which in this case
has the same computational cost 
as kernel $k$-means algorithm but usually with better clustering
quality. We provide carefully designed numerical experiments
showing the superiority of this method compared to
kernel $k$-means, standard $k$-means, and gaussian mixture models.
\end{abstract}

@@

Now we write this in terms of the matrix $Y = Z D^{-1/2}$.
The objective function immediately becomes
$\Tr\left( Y^\top G \, Y\right)$. Notice that the above constraints
imply that $Z^T Z = D$, where $D=\diag(n_1,\dotsc,n_k)$, which in turn gives
$D^{-1/2} Y^T Y D^{-1/2} = D$, or $Y^\top Y = I$. 
Also, every entry of $Y$ is positive by definition,
$Y \ge 0$. Now it only remains to show the last 
constraint in \eqref{eq:qcqp2}, which comes from the last
constraint in \eqref{eq:qcqp}. In matrix form this reads
$Z^T \e = D \e$. Replacing $Z=YD^{1/2}$ we have
$Y^\top \e = D^{1/2} \e$. Multiplying this last equation
on the left by $Y$, and noticing
that $Y D^{1/2} \e = Z \e = \e$, we finally obtain
$Y Y^\top \e = \e$. Therefore, the optimization 
problem \eqref{eq:qcqp} is equivalent
to \eqref{eq:qcqp2} .
\end{proof}

Based on Proposition~\ref{th:qcqp2}, to group data
$\mathbb{X} = \{ x_1,\dotsc,x_n \}$
into  $k$ clusters we first compute the Gram matrix
$G$ and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$. 
Problem \eqref{eq:qcqp2} is NP-hard and there
are few methods
available to solve it directly,
which is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints, or obtaining a relaxed SDP version of it.
For instance, the relaxed problem
\begin{equation}
\max_{Y} \Tr \left( Y^\top G \, Y \right) \quad \mbox{s.t. $Y^\top Y = I$}
\end{equation}
has a well-known closed form solution $Y^\star = U R$, where the
columns of $U$ contain the leading $k$ eigenvectors of $G$ corresponding
to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and
$R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
The resulting
optimal objective function is given by
$\max \Tr \left( {Y^\star}^\top G \, Y^\star \right)  = 
\sum_{i=1}^k \lambda_i$. One might then normalize and threshold the rows
of $Y^\star$, or better, following \cite{NgJordan} one can normalize the
rows of $Y^\star$ and apply standard $k$-means on this matrix where each
row is considered as a data point.
This is the procedure done in spectral clustering on a graph Laplacian
matrix obtained through a similarity matrix.
However, computing eigenvectors of very large matrices
can be prohibitively expensive 
and usually iterative methods are preferred. Moreover,
this procedure is only guaranteed to converge if the normalized $Y^\star$ is
positive semidefinite.
In Section~\ref{sec:algo} we will propose an iterative method to directly 
find approximate solutions to \eqref{eq:qcqp2} that is guaranteed to
converge and does not require that the similarities $\kk(x_i,x_j)$ be
nonnegative.

@@

The clustering problem \eqref{eq:qcqp2} based on energy statistics
is valid for data living in an \emph{arbitrary} space of negative type where
a semimetric $\rho$, and thus the kernel \eqref{eq:kernel_semimetric}, are
assumed to be known. This clustering method is
\emph{nonparametric} since it does not make any assumptions
about  the distribution of the data, 
contrary to $k$-means and GMM, for example.
If one uses the standard energy distance
defined in \eqref{eq:energy} 
we have $\rho(x,y) = \| x - y\|$ and this fixes the kernel
through \eqref{eq:kernel_semimetric}. In the same way, 
for data living in a more general
metric space $(\mathcal{X}, \rho)$ the corresponding semimetric $\rho$ fixes
the kernel.
In practice, however,
the clustering quality strongly depend on the choice of a suitable
$\rho$ which measures the similarity between data points,
and is equivalent to choosing an appropriate kernel.
The energy distance \eqref{eq:energy2} fixes a family of choices
$\rho(x,y) = \| x-y\|^\alpha$ with $0 < \alpha\le 2$. 
Nevertheless, if prior information
is available one may choose a more appropriate $\rho$.

\subsection*{Relation to Kernel $\bm{k}$-Means}
One may wonder how energy statistics clustering 
relates to the well-known kernel $k$-means problem%
\footnote{When we refer to kernel $k$-means problem we mean specifically 
the optimization problem \eqref{eq:kernel_kmeans}, which should not be 
confused with kernel $k$-means algorithm that is just one possible recipe 
to solve \eqref{eq:kernel_kmeans}.
} 
which is extensively used in machine learning.
%and was proposed as a kernelized version of $k$-means.
For a positive semidefinite Gram matrix $G$, as defined in
\eqref{eq:kernel_matrix},
there exists a map
$\varphi: \mathcal{X} \to \HH_\kk$ such that
$\kk(x,y) = \varphi(x)^\top \varphi(y)$. The kernel $k$-means optimization
problem,
in feature space,
is defined by
\begin{equation}
\label{eq:kernel_kmeans}
\min_{\C_1,\dotsc,\C_k}\bigg\{ 
J(\C_1,\dots,\C_k) \equiv  \sum_{j=1}^k
\sum_{x \in \C_j} \| \varphi(x) - \varphi(\mu_j) \|^2
\bigg\}
\end{equation}
where $\mu_j = \tfrac{1}{n_j} \sum_{x \in \C_j} x$ is the  mean of cluster
$\C_j$ in the ambient space. Notice that the above objective function
is strongly tied to the idea of minimizing distances between points
and cluster centers, which arises from $k$-means objective function.
It was shown \cite{Dhillon2,Dhillon} 
that problem \eqref{eq:kernel_kmeans} 
is equivalent to a QCQP in the same form as
\eqref{eq:qcqp2}. The next result makes this explicit, showing that
\eqref{eq:minimize} and \eqref{eq:kernel_kmeans} are actually equivalent.

@@

The above result shows that 
kernel $k$-means problem is actually a consequence
of energy statistics. In this vein, kernel $k$-means is part of a
statistical framework where distances between probability
distributions are maximized. 
The kernel function $\kk$ arises from a semimetric $\rho$ of
negative type which defines the energy distance between distributions.

@@

The relation between kernel $k$-means and graph partitioning problems
is known \cite{Dhillon2,Dhillon}. For conciseness, we repeat a similar 
analysis due to the relation of these problems to
energy statistics and RKHS, which may provide a different perspective.

@@

Consider \eqref{eq:assoc}, whose objective function can be written as
\begin{equation}
\sum_{j=1}^k \dfrac{1}{s_j} \sum_{p \in \C_j} \sum_{q \in \C_j}
\mathcal{A}_{pq} = \sum_{j=1}^k \sum_{p=1}^n \sum_{q=1}^n 
\dfrac{Z^\top_{jp}}{\sqrt{s_j}} \, \mathcal{A}_{pq} \, 
\dfrac{Z_{qj}}{\sqrt{s_j}}
= \Tr\left( Y^\top \mathcal{A} Y \right) ,
\end{equation}
with $Z$ defined in \eqref{eq:label_matrix} and $Y$ in
\eqref{eq:weighted_matrices}. To make the analogy with \eqref{eq:qcqp3}
explicit,  problem \eqref{eq:assoc} is equivalent to
\begin{equation}
\max_H \Tr\left( H^\top \mathcal{W}^{-1/2} \mathcal{A} \mathcal{W}^{-1/2} H 
\right) \qquad \mbox{s.t. $H\ge 0$, $H^\top H = I$, $H H^\top
\om=\om$}.
\end{equation}
Therefore, this is exactly the same problem as weighted energy statistics
clustering \eqref{eq:qcqp3} with 
$G = \mathcal{W}^{-1} \mathcal{A} \mathcal{W}^{-1}$. Assuming this
matrix is positive semidefinite, this generates a semimetric
\eqref{eq:gen_kernel} for graphs given by
\begin{equation}
\label{eq:metric_graphs}
\rho(i,j) = 
\dfrac{\mathcal{A}_{ii}}{w_i^{2}}
+\dfrac{\mathcal{A}_{jj}}{w_j^{2}}
-\dfrac{2 \mathcal{A}_{ij}}{w_i w_j} \qquad\mbox{or}\qquad
\rho(i,j) = -\dfrac{2 \mathcal{A}_{ij}}{w_i w_j}
\end{equation}
for vertices $i,j \in \mathcal{V}$, and 
where in the second equation we assume the graph has no self-loops,
i.e. $\mathcal{A}_{ii} = 0$. Using \eqref{eq:metric_graphs} 
in \eqref{eq:g_def}--\eqref{eq:between}
allows one to use energy statistics theory for inference
on graphs.

@@

For a two-class problem we can use \eqref{eq:w1d} to cluster the data
through a simple algorithm 
as follows. We first order
the entire dataset, $\mathbb{X} \to \bar{\mathbb{X}}$. Then 
we compute \eqref{eq:w1d} for each possible split of $\bar{\mathbb{X}}$
and pick the point which gives the minimum value of $W$. 
This procedure is described in Algorithm~\ref{algo1d}. 
Note that
this method does not require any initialization,
however,
it only works for one-dimensional data with Euclidean distance. The total
complexity of the algorithm is $\OO(n\log n + n^2) = \OO(n^2)$.

@@

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=.8\textwidth]{hist_normal.pdf}\\[-.8em]
\includegraphics[width=\textwidth]{gauss1d.pdf}\\[-1em]
(a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=.8\textwidth]{hist_lognormal.pdf}\\[-.8em]
\includegraphics[width=\textwidth]{loggauss1d.pdf}\\[-1em]
(b)
\end{minipage}
\caption{
\label{fig:1d}
Energy statistics clustering by Algorithm~\ref{algo1d}
compared to $k$-means and GMM/EM.
We have the same number of points in both clusters, and for each case
we sample $100$ times from the distributions shown in the histograms. We 
plot the average value of 
\eqref{eq:accuracy} versus the total number of points 
(error bars are standard error). The dashed line indicates the
best possible classification accuracy computed from Bayes error.
(a) Data coming from \eqref{eq:two_normal}, where the
optimal accuracy is $\approx 0.956$.
(b) Data from \eqref{eq:two_lognormal}, where the optimal
accuracy is $\approx 0.852$.
}
\end{figure}

Let us consider two simple experiments with equal number of points
in each cluster. We plot the accuracy \eqref{eq:accuracy} versus the number
of points in each cluster.
The data is clustered using 
Algorithm~\ref{algo1d}, GMM through EM algorithm, and standard $k$-means 
algorithm. In the last two cases 
we use the initialization from $k$-means++ \cite{Vassilvitskii} 
and we run the algorithms
few times choosing the result with  best objective function value. 
Notice that Algorithm~\ref{algo1d} requires no initialization so we only
run it once.
Moreover, for each case we sample $100$ times
and show the average accuracy with error bars indicating
the standard error.
In  Fig.~\ref{fig:1d}a 
we have data sampled from two normal distributions with equal number of
points in each cluster,
\begin{equation}
\label{eq:two_normal}
x \sim \mathcal{N}\big(\mu_i,\sigma_i^2\big) 
\quad 
\mbox{with $\mu_1 = 0$, $\sigma_1=1$ and
$\mu_2 = 5$, $\sigma_2 = 2$.}
\end{equation}
For these distributions the optimal accuracy 
obtained from Bayes classification error
is $ \approx 0.956$ which is indicated by the dashed line in the plot.
We see that the three methods
perform closely. As expected, GMM has a slight advantage over the other
methods since
it corresponds to the true model of the data. 
Energy statistics performs slightly better
than $k$-means. On the other hand, in Fig.~\ref{fig:1d}b
we consider two clusters with lognormal distributions,
\begin{equation}
\label{eq:two_lognormal}
\log x \sim \mathcal{N}\big( \mu_i,\sigma_i^2\big) \quad
\mbox{with
$\mu_1 = 0$,
$\sigma_1 = 0.3$ and
$\mu_2 = -1.5$,
$\sigma_2 = 1.5$.}
\end{equation}
The optimal classification accuracy from Bayes error is $\approx 0.852$.
In this case,
Algorithm~\ref{algo1d} provides a very accurate clustering while
GMM and $k$-means basically cluster at chance.
Sometimes GMM/EM  was unable to estimate the parameters 
thus giving zero accuracy.
The two simple experiments of Fig.~\ref{fig:1d} illustrate
how energy statistics clustering is nonparametric, being able
to provide high quality clustering in settings where data comes
from very different distributions.

@@

In this section we introduce an iterative algorithm to find a local
maximizer of \eqref{eq:qcqp2}. However, due to 
Proposition~\ref{th:kernel_kmeans} we can also find an approximate
solution by the well-known kernel $k$-means algorithm 
(see \cite{Dhillon2,Dhillon}), which 
for convenience will also be restated in the present context.

@@

To optimize the kernel $k$-means objective function
\eqref{eq:J} we remove the global term and define the function
\begin{equation}
\label{eq:Jell}
J^{(\ell)}(x_i) \equiv 
\dfrac{1}{n_\ell^2} Q_\ell
-\dfrac{2}{n_\ell} Q_\ell(x_i) 
\end{equation}
which represents a cost depending on point $x_i$ and cluster $\C_\ell$. One
thus assigns  $x_i$ to cluster $\C_{j^\star}$ according
to $j^\star = \argmin_\ell J^{(\ell)}(x_i)$, for $\ell = 1,\dotsc,k$.
This procedure is performed for every data point and repeated until
convergence, i.e. until no new assignments are made.
The entire procedure is described in Algorithm~\ref{kmeans_algo}.
It can be shown that this algorithm converges provided $G$ is positive
semidefinite.

@@

Let us mention some important results about Hartigan's method.

@@

Both conditions of the  above theorem are not satisfied 
by Lloyd's method, and consequently by (kernel) $k$-means algorithm.
The next result indicates that Hartigan's can potentially 
escape local optima of Lloyd's method.

@@

The main goal of the 
experiments  to follow
is to compare Algorithm~\ref{algo} based on Hartigan's method to
kernel $k$-means, as described in Algorithm~\ref{kmeans_algo}, which
is based on Lloyd's method. 
From the discussion in the end of the previous section we can anticipate
that Algorithm~\ref{algo} will be superior than Algorithm~\ref{kmeans_algo}.
Another goal is to illustrate the nonparametric aspect of energy statistics.
To this end we also compare
Algorithm~\ref{algo} to standard $k$-means and GMM/EM since 
these are the most used
clustering algorithms in practice.
Moreover, for every algorithm, we always
choose the initialization procedure from $k$-means++%
\footnote{Notice that we just use the initialization
procedure and not the full $k$-means++ algorithm.} \cite{Vassilvitskii}.
Our measure of clustering quality will be the accuracy \eqref{eq:accuracy}
based on the ground truth. Furthermore, in every experiment, we sample
data many times and show the average value of the accuracy with error bars
indicating the standard error. Whenever possible, 
we also indicate the optimal 
accuracy computed from
Bayes classification error.

>>>>

\begin{abstract}
Energy statistics 
was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics,
and it provides a nonparametric test for equality of distributions.
It was generalized to 
metric spaces of strong negative type,
and more recently a connection with reproducing kernel Hilbert spaces 
was proposed.
Although  
extensively used in statistics, it
has not been previously incorporated in machine learning problems.
% @gui: not sure what you mean?  what counts as machine learning?  be specific, i think yo mean clustering?  even then, it is not true, both the master's thesis and the RKHS paper kind of do it.  be as general as possible about the gap that we are filling, but not more general.
Here we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP), which we show
to be equivalent
to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of 
kernel $k$-means from energy statistics.
% @gui: have other people shown a relationship between kernel k-means and say, RKHS, or something else?  
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, which makes connection to 
graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative
algorithm based on Hartigan's method, which in this case
has the same computational cost 
as kernel $k$-means algorithm but usually with better clustering
quality. We provide carefully designed numerical experiments
showing the superiority of this method compared to
kernel $k$-means, standard $k$-means, and gaussian mixture models in a variety of settings.
\end{abstract}

@@

Now we write this in terms of the matrix $Y = Z D^{-1/2}$.
% @gui: Y is a matrix? or a vector? when introducing new terms, always define what space they are in rigorously
The objective function immediately becomes
$\Tr\left( Y^\top G \, Y\right)$. Notice that the above constraints
imply that $Z^T Z = D$, where $D=\diag(n_1,\dotsc,n_k)$, which in turn gives
$D^{-1/2} Y^T Y D^{-1/2} = D$, or $Y^\top Y = I$. 
Also, every entry of $Y$ is positive by definition,
$Y \ge 0$. Now it only remains to show the last 
constraint in \eqref{eq:qcqp2}, which comes from the last
constraint in \eqref{eq:qcqp}. In matrix form this reads
$Z^T \e = D \e$. Replacing $Z=YD^{1/2}$ we have
$Y^\top \e = D^{1/2} \e$. Multiplying this last equation
on the left by $Y$, and noticing
that $Y D^{1/2} \e = Z \e = \e$, we finally obtain
$Y Y^\top \e = \e$. Therefore, the optimization 
problem \eqref{eq:qcqp} is equivalent
to \eqref{eq:qcqp2} .
% @gui: don't use equation numbers here.
\end{proof}

% @jv: stopped here
Based on Proposition~\ref{th:qcqp2}, to group data
$\mathbb{X} = \{ x_1,\dotsc,x_n \}$
into  $k$ clusters we first compute the Gram matrix
$G$ and then 
solve the optimization problem \eqref{eq:qcqp2} for $\Zt \in
\mathbb{R}^{n\times k}$. The $i$th row
of $\Zt$ will contain a single nonzero element in some $j$th column,
indicating that $x_i \in \C_j$. 
Problem \eqref{eq:qcqp2} is NP-hard and there
are few methods
available to solve it directly,
which is computational prohibitive even for small datasets.
However, one can find approximate solutions by relaxing some 
of the constraints, or obtaining a relaxed SDP version of it.
For instance, the relaxed problem
\begin{equation}
\max_{Y} \Tr \left( Y^\top G \, Y \right) \quad \mbox{s.t. $Y^\top Y = I$}
\end{equation}
has a well-known closed form solution $Y^\star = U R$, where the
columns of $U$ contain the leading $k$ eigenvectors of $G$ corresponding
to the $k$ largest eigenvalues $\lambda_1\ge \lambda_2\ge\dotsc\ge\lambda_k$, 
and
$R \in \mathbb{R}^{k\times k}$ is an arbitrary orthogonal matrix. 
The resulting
optimal objective function is given by
$\max \Tr \left( {Y^\star}^\top G \, Y^\star \right)  = 
\sum_{i=1}^k \lambda_i$. One might then normalize and threshold the rows
of $Y^\star$, or better, following \cite{NgJordan} one can normalize the
rows of $Y^\star$ and apply standard $k$-means on this matrix where each
row is considered as a data point.
This is the procedure done in spectral clustering on a graph Laplacian
matrix obtained through a similarity matrix.
However, computing eigenvectors of very large matrices
can be prohibitively expensive 
and usually iterative methods are preferred. Moreover,
this procedure is only guaranteed to converge if the normalized $Y^\star$ is
positive semidefinite.
In Section~\ref{sec:algo} we will propose an iterative method to directly 
find approximate solutions to \eqref{eq:qcqp2} that is guaranteed to
converge and does not require that the similarities $\kk(x_i,x_j)$ be
nonnegative.

@@

The clustering problem \eqref{eq:qcqp2} based on energy statistics
is valid for data living in an \emph{arbitrary} space of negative type where
a semimetric $\rho$, and thus the kernel \eqref{eq:kernel_semimetric}, are
assumed to be known. This clustering method is
\emph{nonparametric} since it does not make any assumptions
about  the distribution of the data, 
contrary to $k$-means and GMM, for example.
% @gui: algorithms don't make assumptions, proofs do.  we don't prove anything about when we get the right answer, so we can't quite say this.
If one uses the standard energy distance
defined in \eqref{eq:energy} 
we have $\rho(x,y) = \| x - y\|$ and this fixes the kernel
through \eqref{eq:kernel_semimetric}. In the same way, 
for data living in a more general
metric space $(\mathcal{X}, \rho)$ the corresponding semimetric $\rho$ fixes
the kernel.
In practice, however,
the clustering quality strongly depend on the choice of a suitable
$\rho$ which measures the similarity between data points,
and is equivalent to choosing an appropriate kernel.
The energy distance \eqref{eq:energy2} fixes a family of choices
$\rho(x,y) = \| x-y\|^\alpha$ with $0 < \alpha\le 2$. 
Nevertheless, if prior information
is available one may choose a more appropriate $\rho$.

\subsection*{Relation to Kernel $\bm{k}$-Means}
One may wonder how energy statistics clustering 
relates to the well-known kernel $k$-means problem%
\footnote{When we refer to kernel $k$-means problem we mean specifically 
the optimization problem \eqref{eq:kernel_kmeans}, which should not be 
confused with kernel $k$-means algorithm that is just one possible recipe 
to solve \eqref{eq:kernel_kmeans}.
} 
which is extensively used in machine learning.
%and was proposed as a kernelized version of $k$-means.
For a positive semidefinite Gram matrix $G$, as defined in
\eqref{eq:kernel_matrix},
there exists a map
$\varphi: \mathcal{X} \to \HH_\kk$ such that
$\kk(x,y) = \varphi(x)^\top \varphi(y)$. The kernel $k$-means optimization
problem,
in feature space,
is defined by
\begin{equation}
\label{eq:kernel_kmeans}
\min_{\C_1,\dotsc,\C_k}\bigg\{ 
J(\C_1,\dots,\C_k) \equiv  \sum_{j=1}^k
\sum_{x \in \C_j} \| \varphi(x) - \varphi(\mu_j) \|^2
\bigg\}
\end{equation}
where $\mu_j = \tfrac{1}{n_j} \sum_{x \in \C_j} x$ is the  mean of cluster
$\C_j$ in the ambient space. Notice that the above objective function
is strongly tied to the idea of minimizing distances between points
and cluster centers, which arises from $k$-means objective function.
 \cite{Dhillon2,Dhillon} show
that problem \eqref{eq:kernel_kmeans} 
is equivalent to a QCQP in the same form as
\eqref{eq:qcqp2}. The next result makes this explicit, showing that
\eqref{eq:minimize} and \eqref{eq:kernel_kmeans} are actually equivalent.

@@

The above result shows that 
kernel $k$-means problem is actually a consequence
% @gui: i'm not sure about this.  who is to say the correct direction of causality? maybe energy clustering is a special case of kernel k-means?
of energy statistics. In this vein, kernel $k$-means is part of a
statistical framework where distances between probability
distributions are maximized. 
The kernel function $\kk$ arises from a semimetric $\rho$ of
negative type which defines the energy distance between distributions.

@@

The relation between kernel $k$-means and graph partitioning problems
is known \cite{Dhillon2,Dhillon}. For conciseness, we repeat a similar 
analysis due to the relation of these problems to
energy statistics and RKHS, which  provides a different perspective.

@@

Consider \eqref{eq:assoc}, whose objective function can be written as
\begin{equation}
\sum_{j=1}^k \dfrac{1}{s_j} \sum_{p \in \C_j} \sum_{q \in \C_j}
\mathcal{A}_{pq} = \sum_{j=1}^k \sum_{p=1}^n \sum_{q=1}^n 
\dfrac{Z^\top_{jp}}{\sqrt{s_j}} \, \mathcal{A}_{pq} \, 
\dfrac{Z_{qj}}{\sqrt{s_j}}
= \Tr\left( Y^\top \mathcal{A} Y \right) ,
\end{equation}
with $Z$ defined in \eqref{eq:label_matrix} and $Y$ in
\eqref{eq:weighted_matrices}. To make the analogy with \eqref{eq:qcqp3}
explicit,  problem \eqref{eq:assoc} is equivalent to
\begin{equation}
\max_H \Tr\left( H^\top \mathcal{W}^{-1/2} \mathcal{A} \mathcal{W}^{-1/2} H 
\right) \qquad \mbox{s.t. $H\ge 0$, $H^\top H = I$, $H H^\top
\om=\om$}.
\end{equation}
Therefore, this is exactly the same problem as weighted energy statistics
clustering \eqref{eq:qcqp3} with 
$G = \mathcal{W}^{-1} \mathcal{A} \mathcal{W}^{-1}$. Assuming this
matrix is positive semidefinite, this generates a semimetric
\eqref{eq:gen_kernel} for graphs given by
\begin{equation}
\label{eq:metric_graphs}
\rho(i,j) = 
\dfrac{\mathcal{A}_{ii}}{w_i^{2}}
+\dfrac{\mathcal{A}_{jj}}{w_j^{2}}
-\dfrac{2 \mathcal{A}_{ij}}{w_i w_j} \qquad\mbox{or}\qquad
\rho(i,j) = -\dfrac{2 \mathcal{A}_{ij}}{w_i w_j}
\end{equation}
for vertices $i,j \in \mathcal{V}$, and 
where in the second equation we assume the graph has no self-loops,
i.e. $\mathcal{A}_{ii} = 0$. Using \eqref{eq:metric_graphs} 
in \eqref{eq:g_def}--\eqref{eq:between}
allows one to use energy statistics theory for inference
on graphs.
% @gui: can you provide some guidance here, in particular, what do the weights "mean" or "correspond to"? right now, it feels like just math.

@@

For a two-class problem we can use \eqref{eq:w1d} to cluster the data
through a simple algorithm 
as follows. We first order
the entire dataset, $\mathbb{X} \to \bar{\mathbb{X}}$. Then 
we compute \eqref{eq:w1d} for each possible split of $\bar{\mathbb{X}}$
and pick the point which gives the minimum value of $W$. 
This procedure is described in Algorithm~\ref{algo1d}. 
Note that
this method does not require any initialization,
however,
it only works for one-dimensional data with Euclidean distance. The total
complexity of the algorithm is $\OO(n\log n + n^2) = \OO(n^2)$.
% @gui: "does not require any initiializaton" i think you mean is guaranteed to find the optimal solution in O(n^2) time.

@@

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=.8\textwidth]{hist_normal.pdf}\\[-.8em]
\includegraphics[width=\textwidth]{gauss1d.pdf}\\[-1em]
(a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=.8\textwidth]{hist_lognormal.pdf}\\[-.8em]
\includegraphics[width=\textwidth]{loggauss1d.pdf}\\[-1em]
(b)
\end{minipage}
\caption{
% @gui: to provide intuition to the reader, let's show the kernel density estimate for each of the 3 algorithms for both classes.  i'm not sure the best way to visualize it, maybe play with a few options.
% @gui: also, the colors are not very different.  i use the following color pallete:
%co = [0,255,0; % green
%0,255,255;  % cyan
%255,0,255;  % magenta
%255,127,0;  % orange
%55,126,184; % blue
%153,153,153;% gray
%152,78,163; % purple
%166,86,40;  % brown
%247,129,191;% pink
%77,175,74;  % forest
%228,26,28;  % red
%255,255,51];% yellow
\label{fig:1d}
Energy statistics clustering by Algorithm~\ref{algo1d}
compared to $k$-means and GMM/EM.
We have the same number of points in both clusters, and for each case
we sample $100$ times from the distributions shown in the histograms. We 
plot the average value of 
\eqref{eq:accuracy} versus the total number of points 
(error bars are standard error). The dashed line indicates the
best possible classification accuracy computed from Bayes error.
(a) Data coming from \eqref{eq:two_normal}, where the
optimal accuracy is $\approx 0.956$.
(b) Data from \eqref{eq:two_lognormal}, where the optimal
accuracy is $\approx 0.852$.
}
\end{figure}

Let us consider two simple experiments with equal number of points
in each cluster. We plot the accuracy \eqref{eq:accuracy} versus the number
of points in each cluster.
The data is clustered using 
Algorithm~\ref{algo1d}, 
% @gui: i see why you call this algorithm 1, but can we call it something with semantic meaning? like 1D-\mathcal{E}-clustering? or maybe \mathcal{E}1 for short? same for the other variants.
GMM through EM algorithm, and standard $k$-means 
algorithm. 
% @gui: there are many k-means algorithms, be specific.
In the last two cases 
we use the initialization from $k$-means++ \cite{Vassilvitskii} 
and we run the algorithms
few times choosing the result with  best objective function value. 
% @gui: a few? be specific
Notice that Algorithm~\ref{algo1d} requires no initialization so we only
% @gui: "no initiazation" is weird.  do any algs require an initialization?
run it once.
Moreover, for each case we sample $100$ times
and show the average accuracy with error bars indicating
the standard error.
In  Fig.~\ref{fig:1d}a 
we have data sampled from two normal distributions with equal number of
points in each cluster,
\begin{equation}
\label{eq:two_normal}
x \sim \mathcal{N}\big(\mu_i,\sigma_i^2\big) 
\quad 
\mbox{with $\mu_1 = 0$, $\sigma_1=1$ and
$\mu_2 = 5$, $\sigma_2 = 2$.}
\end{equation}
For these distributions the optimal accuracy 
obtained from Bayes classification error
is $ \approx 0.956$ which is indicated by the dashed line in the plot.
We see that the three methods
perform closely. As expected, GMM has a slight advantage over the other
methods since
it corresponds to the true model of the data. 
Energy statistics performs slightly better
than $k$-means. On the other hand, in Fig.~\ref{fig:1d}b
we consider two clusters with lognormal distributions,
\begin{equation}
\label{eq:two_lognormal}
\log x \sim \mathcal{N}\big( \mu_i,\sigma_i^2\big) \quad
\mbox{with
$\mu_1 = 0$,
$\sigma_1 = 0.3$ and
$\mu_2 = -1.5$,
$\sigma_2 = 1.5$.}
\end{equation}
% @gui: why not use the same parameters as in fig 1a?  i feel like if we did, it would seem less like we are cherry picking?  
The optimal classification accuracy from Bayes error is $\approx 0.852$.
% @gui: computed numerically?
In this case,
Algorithm~\ref{algo1d} provides a very accurate clustering while
GMM and $k$-means basically cluster at chance.
Sometimes GMM/EM  was unable to estimate the parameters 
thus giving zero accuracy.
The two simple experiments of Fig.~\ref{fig:1d} illustrate
how energy statistics clustering is nonparametric, being able
to provide high quality clustering in settings where data comes
from very different distributions.

@@

In this section we introduce an iterative algorithm to find a local
maximizer of \eqref{eq:qcqp2}. Due to 
Proposition~\ref{th:kernel_kmeans} we can also find an approximate
solution by the well-known kernel $k$-means algorithm 
(see \cite{Dhillon2,Dhillon}), which 
for convenience will also be restated in the present context.

@@

To optimize the kernel $k$-means objective function
\eqref{eq:J} we remove the global term and define the function
\begin{equation}
\label{eq:Jell}
J^{(\ell)}(x_i) \equiv 
\dfrac{1}{n_\ell^2} Q_\ell
-\dfrac{2}{n_\ell} Q_\ell(x_i) 
\end{equation}
which represents a cost depending on point $x_i$ and cluster $\C_\ell$. One
thus assigns  $x_i$ to cluster $\C_{j^\star}$ according
to $j^\star = \argmin_\ell J^{(\ell)}(x_i)$, for $\ell = 1,\dotsc,k$.
This procedure is performed for every data point and repeated until
convergence, i.e. until no new assignments are made.
The entire procedure is described in Algorithm~\ref{kmeans_algo}.
% @gui: let's call this something more informative
It can be shown that this algorithm converges provided $G$ is positive
semidefinite.

@@

Let us mention some important results about Hartigan's method.
% @gui: "let us" again...

@@

Both conditions of the  above theorem are not satisfied 
by Lloyd's method, 
% @gui: do you mean Lloyd's method satisfies neither, or does not satisfy both?  please clarify
and consequently by (kernel) $k$-means algorithm.
The next result indicates that Hartigan's can potentially 
escape local optima of Lloyd's method.

@@

The main goal of the 
experiments  to follow
is to compare Algorithm~\ref{algo} based on Hartigan's method to
kernel $k$-means, as described in Algorithm~\ref{kmeans_algo}, which
is based on Lloyd's method. 
From the discussion in the end of the previous section we can anticipate
that Algorithm~\ref{algo} will be superior than Algorithm~\ref{kmeans_algo}.
Another goal is to illustrate the nonparametric aspect of energy statistics.
To this end we also compare
Algorithm~\ref{algo} to standard $k$-means and GMM/EM since 
these are reference
clustering algorithms in practice.
Moreover, for every algorithm, we always
choose the initialization procedure from $k$-means++%
\footnote{Notice that we just use the initialization
procedure and not the full $k$-means++ algorithm.} \cite{Vassilvitskii}.
Our measure of clustering quality will be the accuracy \eqref{eq:accuracy}
based on the ground truth. Furthermore, in every experiment, we sample
data many times and show the average value of the accuracy with error bars
indicating the standard error. Whenever possible, 
% @gui: bayes error is always possible to numerically estimate, no? maybe sometimes computationally intractable?
we also indicate the optimal 
accuracy computed from
Bayes classification error.
